---
title: "Data Analyses of riverNarva in two sampling spots: Vasknarva and Narva"
author: "Kati Roosalu"
date: "18.12.2019"
output: html_notebook
---



#### Read in Vasknarva and Narva sampling spot data from excel fail (read_xlsx)

```{r}
vasknarva<-read_xlsx('Andmed/ID2Vasknarva.xlsx')
narva<-read_xlsx('Andmed/ID3Narva.xlsx')
```
#### View the datatables
```{r}
View(vasknarva)
View(narva)
```
#### Show first 3 rows of data table
```{r}
head(vasknarva,3)
```

#### Data: in the format of excel fails

* Date: Time of water sample taken; yyyy/mm/dd (date)
* TotN: Concentration of total nitrogen; mg/l (numeric)
* TotP: Concentration of total phosphorous; mg/l (numeric)
* NvsP: TotN and TotP ratio (N/P) (numeric)
* ID: Rivers ID (numeric)
* Season - Different time periods: 1) Winter-month from Nov-Apr; 2) Veg.per-month from May-Oct (character)
* Cloroa: amount of chlorphyll, microg/l (numeric)


#### Make plot of TotN vs time in Vasknarva. Add line to show trend (stat_smooth).
```{r}
spVasknarvaN<-ggplot(vasknarva, aes(x=Date, y=TotN))+
  geom_line(color='red')+
  geom_point(color='darkred')+
  labs(x='Years',y='TotN; mg/l')+
  theme_classic()+
  ggtitle('Vasknarva, Total N; mg/l, 2008-2018')+
  stat_smooth(method = "lm", col = "firebrick", se = F, size = 1.3)
spVasknarvaN
```

#### Make plot of TotP vs time in Vasknarva. Add line to show trend.
```{r}
spVasknarvaP<-ggplot(vasknarva, aes(x=Date, y=TotP))+
  geom_line(color='green')+
  geom_point(color='darkgreen')+
  labs(x='Years',y='TotP; mg/l')+
  theme_classic()+
  ggtitle('Vasknarva, Total P; mg/l, 2008-2018')+
  stat_smooth(method = "lm", col = "firebrick", se = F, size = 1.3)
spVasknarvaP
```

#### Make plot of TotN vs time in Narva. Add line to show trend.
```{r}
spNarvaN<-ggplot(narva, aes(x=Date, y=TotN))+
  geom_line(color='red')+
  geom_point(color='darkred')+
  labs(x='Years',y='TotN; mg/l')+
  theme_classic()+
  ggtitle('Narva, Total N; mg/l, 2008-2018')+
  stat_smooth(method = "lm", col = "firebrick", se = F, size = 1.3)
spNarvaN
```

#### Make plot of TotP vs time in Narva. Add line to show trend.

```{r}
spNarvaP<-ggplot(narva, aes(x=Date, y=TotP))+
  geom_line(color='green')+
  geom_point(color='darkgreen')+
  labs(x='Years',y='TotP; mg/l')+
  theme_classic()+
  ggtitle('Narva, Total P; mg/l, 2008-2018')+
  stat_smooth(method = "lm", col = "firebrick", se = F, size = 1.3)
spNarvaP
```


#### For comparing two sampling spots, place plots of N next to eachother (ggarrange). Remove y axis title for Narva (rremove).
```{r}
koosN<-ggarrange(spVasknarvaN,spNarvaN+rremove('ylab'))
koosN
```

#### Place plots of P next to eachother.

```{r}
koosP<-ggarrange(spVasknarvaP,spNarvaP+rremove('ylab'))
koosP
```

#### Add N:P ratio to Vasknarva 
```{r}
NPVask<-ggplot(vasknarva,aes(x=TotN,y=TotP,color=NvsP))+
  geom_point()+
  labs(x='Total N; mg/l', y='Total P; mg/l')+
  ggtitle('N and P concentrations in Vasknarva, 2008-2018')+
  scale_color_gradient(name='N:P ratio',breaks=c(10,20,50),low='red',high='green')+
  theme_light() 
NPVask
```

#### Add new variable 'Season'-> Veg.per and Winter to Vasknarva
```{r}
gpNarva1<-ggplot(vasknarva, aes(x=Date, y=TotN, color=Season))+geom_point()+
ggtitle('Total nitrogen concentrations in Vasknarva, 2008-2018')+
  labs(x='Years', y='TotN; mg/l')+
  scale_color_manual('Seasons:', values=c('green','blue'),
                     labels=c('Veg.per (May-Oct)', 'Winter (Nov-Apr)'))+
  theme(plot.title=element_text(hjust=0.5))+
  theme_light()
gpNarva1
```

#### Add mean value line to two 'Season' groups (stat_smooth) to see if there is a difference in mean concentration in two seasons.
```{r}
ggplot(vasknarva, aes(x=Date, y=TotN, color=Season))+
  geom_point()+
ggtitle('Total nitrogen concentrations in Vasknarva, 2008-2018')+
  labs(x='Years', y='TotN; mg/l')+
  scale_color_manual('Seasons:', values=c('green','blue'),
                     labels=c('Veg.per (May-Oct)', 'Winter (Nov-Apr)'))+
  theme(plot.title=element_text(hjust=0.5))+
  theme_light()+
  stat_smooth(method="lm", formula=y~1, se=FALSE)
```

#### For comparison the difference between two seasons is much bigger in river Emajõgi
```{r}
emajogi<-read_xlsx(('Andmed/ID1E.xlsx'))
emajN<-ggplot(emajogi, aes(x = Date, y = TotN, color = Season)) +geom_point()+
   theme_light()+
  labs(x='Years', y='TotN; mg/l')+
  ggtitle('Total nitrogen concentrations in Emajõgi, 2008-2018')+
  scale_color_manual('Seasons:', values=c('green','blue'),
                     labels=c('Veg.per (May-Oct)', 'Winter (Nov-Apr)'))+
stat_smooth(method="lm", formula=y~1, se=FALSE)
emajN
```

#### Add chlorophyl data to Vasknarva (chlorophyl as color, N:P as size)
```{r}
ggNP1<-ggplot(vasknarva,aes(x=TotN,y=TotP,size=NvsP,color=Cloroa))+
  geom_point()+
  theme_light()+
  labs(x='Total N; mg/l', y='Total P; mg/l')+
  ggtitle('Chlorophyl in Vasknarva, 2008-2018')+
  scale_size_continuous(name='N:P ratio',breaks = c(10,20,50,100))
ggNP1
```

#### The grey dots show times when chlorophyl is not analyzed. Remove data without chlorphyl (subset)

```{r}
ggNP2<-ggplot(subset(vasknarva, !is.na(Cloroa)),
       aes(x=TotN,y=TotP,size=NvsP,color=Cloroa))+
  geom_point()+
  labs(x='Total N; mg/l', y='Total P; mg/l')+
  ggtitle('Chlorophyl in Vasknarva, 2008-2018')+
  scale_size_continuous(name='N:P ratio',breaks = c(10,20,50))+
    theme_light()
ggNP2
```


#### Make better colorscale for cloro-a
```{r}
ggplot(subset(vasknarva, !is.na(Cloroa)),aes(x=TotN,y=TotP))+
  geom_point(aes(size=NvsP,color=as.factor(CloroaG)))+
  labs(x='Total N; mg/l', y='Total P; mg/l')+
  ggtitle('Chlorophyl in Vasknarva, 2008-2018')+
  scale_size_continuous(name='N:P ratio',breaks = c(10,20,50))+
  scale_color_discrete(name='Chlorphyl',labels=c('<5','5-10','>10'))+
    theme_light()
```